Brownian motion of main-belt asteroids on human timescales

### 2025-02-20 main.tex
draft finished

\newcommand{\matA}{A}
\newcommand{\matB}{B}

\newcommand{\vcirc}{v_c}
\newcommand{\vrel}{v_{\rm r}}
\newcommand{\Msun}{M_\odot}
\newcommand{\covm}{C}
\newcommand{\lop}{\varpi}
\newcommand{\Lhat}{\hat L}
\usepackage{natbib} 
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
%\usepackage{lineno}
%\linenumbers

@@

High-precision measurements of positions of Solar System bodies have
great value to multiple scientific pursuits.  They can reveal
gravitational accelerations due to undiscovered bodies such as a
hypothetical ``Planet X'' or passing primordial black holes, determine
the masses of known bodies, tests the laws of gravity themselves, and characterize 
the non-gravitational forces that drive many forms of planetary
migration, \eg\ the maintainence of the near-Earth asteroid (NEA)
population.  While most applications of precision orbit determination
have used the 8 major planets as test particles, there are now $>10^6$
known minor planets available for this task, whose astrometric
information content is growing extremely rapidly through large-scale
survey projects such as Gaia, PanSTARRS, and particularly the upcoming
\textit{Legacy Survey of Space and Time} (LSST), which can produce
milliarcsecond-scale accuracy for magnitude-limited small-body populations.  The great majority
of currently-tracked objects are main-belt asteroids (MBAs), so we
wish to investigate the potential power of MBAs as gravitational test
bodies.

The most advanced ephemerides for the Solar System consider the
gravitational forces emanating from the Sun, from the major planets and their large
moons, from up to $\approx300$ of the largest individual MBAs, and
from distributed mass annuli representing the smaller members of
various small-body populations (asteroid belts, Kuiper belt, etc.).
Here we ask: \textit{what is the typical RMS deviation in angular or radial
position that a typical MBA accrues because of its gravitational
encounters with other individual MBAs?}  We are interested not in the
perturbations that can be well described by a multipolar ring model
for the full population; rather the Brownian motion that accumulates
from closer encounters either with MBAs that \emph{are not included} in the
ephemeris model at all.  For those larger asteroids that \emph{are} included
in an ephemeris model, we need to consider the uncertainties in
orbital motion that accrue from the uncertainty $\sigma_M$ in their
masses.  Any such departures became a source of noise in any
inferences that we make from use of these MBAs as
dynamical tracer particles.

While long-term diffusion of orbital elements has been extensively
investigated in the context of planetary-system formation and
planetary migration, we address here the more limited question of how
much the MBAs will have their orbit elements and positions altered
by Brownian motion over a time period $T \lesssim100$~yr of human
observation, in the present dynamical environment.  We will make use
of the currently known population of MBAs, correcting for
incompleteness of current surveys at lower masses---or more precisely,
at fainter absolute magnitudes $H,$ since only a handful of small
asteroids have usefully accurate mass measures.

\section{Calculation overview}
We aim to calculate the Brownian motion variance to an accuracy of
tens of percent.  Greater accuracy in the calculation is not warranted
since one of the principal inputs---the number of MBAs vs mass---is
uncertain to at least this level because of unknowns in the conversion
from $H$ to mass.  We will therefore be at liberty to drop terms that
modify the perturbations by $O(e^2)$ of the tracer asteroid that is
being tracked.    We will assume throughout that the joint
distribution of mass (or $H$) and orbital elements is separable into a
mass distribution and an orbital distribution.  In this scenario, the
Brownian motion is independent of the tracer mass, and involves an
integral over the mass distribution of the deflecting bodies.

@@

All distances will be given in units of the original semimajor axis $a_0$, and all velocities in units of the circular velocity $\vcirc \equiv \sqrt{G\Msun/a_0}.$  In these units, $G\Msun=1,$ the period of the initial orbit is $2\pi$ and the (unperturbed) mean anomaly is $M=t+M_0.$

@@

One of our tasks will be to derive, from the known asteroid population, the rate (per target) of encounters vs the imparted impulse,
\begin{equation}
  \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z},
  \label{eq:dN}
\end{equation}
where the components of $\vecI$ are given in the cylindrical basis
vectors about the asteroid's position at the impulse.
We will estimate this function from the known population,
approximating it as constant within each of three subsets of tracer MBAs:
the inner, middle, and outer populations bounded by the 4:1, 3:1,
5:2, and 2:1 mean-motion resonances with Jupiter.
We will further assume that the impulses on a given target are drawn
independently from this distribution, \ie\ a Poisson process defined
by this rate.  In this case, the only properties of the impulse
distribution that we need are its second moments for components $\alpha \in \{r,\phi,z\}$
\begin{equation}
  \left \langle n I_\alpha^2 \right\rangle \equiv \int d^2I \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z} I_\alpha^2.
\label{eq:nvsq}
\end{equation}
The average of the cross terms $I_rI_z$ and $I_\phi, I_z$ will vanish if the deflector distribution is symmetric in inclination, and we find numerically that the mean $I_rI_\phi$ is small enough to ignore.

From this knowledge of the impulse distribution, our goal is to obtain the covariance matrix of the deviations in the target's observed position, relative to the initial orbit, after some time $T.$ The position observables are the range, plus the latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\theta=z/r,\phi).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible realizations of the impulse history, and the mean anomaly $M$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the deviations of the orbital elements induced by the impulses.  The $\vecq$ components will be selected to be equal to zero before any impulses are applied, and each responds linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $t$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(t,t_i) = \matA(e_0, t, t_i) \cdot \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\vecq(t)$ will be the sum of the $\Delta\vecq(t_i)$ imparted by all impulses applied at times $0<t_i<t.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(t) \equiv \left\langle \vecq(t) \vecq^T(t) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,t) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e,t,t_i) \matA_{k\gamma}(e,t,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e,t,t_i) \matA_{k\gamma}(e,t,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the element shifts $\vecq$ into the observed displacements $\Delta\vecp$.  In linear perturbation theory this will again be expressible as a matrix
\begin{equation}
  \Delta\vecp(e_0,t) = \matB(e_0,t) \cdot \vecq(t).
\label{eq:B}
\end{equation}
Combining this with \eqq{eq:Cp} and \eqq{eq:Cqjk} yields the desired result
\begin{eqnarray}
  \covm^x_{\alpha\beta}(t,e_0) & = & \sum_{jk} B_{\alpha j}(e,t) B_{\beta k}(e,t) \covm^q_{jk}(t) \nonumber \\
  & = & \sum_\gamma \left\langle nI_\gamma^2\right\rangle \int dt_i \left[AB(e_0,t,t_i)\right]_{\alpha\gamma}  \left[AB(e_0,t,t_i)\right]_{\beta\gamma} 
\label{eq:ABq}
\end{eqnarray}
which we would wish to average over the phase of the initial mean anomaly $M_0$ of the MBA.

Section~\ref{sec:impulse} describes the estimation of $\left\langle
  nI_\alpha^2\right\rangle$ for MBA regions.
Section~\ref{sec:propagation} derives the forms of $\matA(e,t,t_i),$
and $\matB(e,t).$  Section~\ref{sec:results} combines these and
summarizes the results and their implications.

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                              \frac{dn}{dH\,dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int db_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{\sigma_M(H)}{bv}\right)^2 \\
  & = & \int dH \frac{dN}{dH} \sigma_M(H)^2  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        2\pi \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be determined from fitting the data of the tracer.
Fitting an ephemeris that includes all of these Individually
detectable encounters is entirely feasible, as long as the relevant
MBAs are known, with modestly accurate orbits.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$

\subsection{Numerical results}
\eqq{eq:nI2} reduces the calculation to a double integral over the
deflector $H$ distribution and the relative velocity $v.$  The
constituents of this calculation are estimated as follows.

@@

Using a simple leapfrog integrator with gravity from the Sun and the 8
major planets, we advance all MBA's orbits from their heliocentric
osculating elements at the epochs given by the MPC, to barycentric
state vectors on 1 May 2025.  Using $kd$-trees to accelerate
pair-finding, we locate all tracer-deflector pairs
that approach within 0.03~AU of each during the $\pm1$-day period
around this initial epoch, assuming inertial relative motion during
this interval.  The circumstances of such encounters are saved: the
identities of the two MBAs involved, the time of closest approach, the
relative velocity $v,$ and the impact parameter vector $\vecb.$

@@

\subsubsection{Asteroid properties}
We crudely approximate the population as having a geometric albedo of 0.25,
density of 2500~kg~m$^{-3},$ and spherical shapes, which leads to the
conversion
\begin{equation}
  M(H) = 1.2\times10^{-17} \times 10^{-0.6(H-15)}\, M_\odot.
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{mbacounts.pdf}
  \caption{Differential counts of MBAs cataloged by the MPC are
    plotted vs absolute magnitude $H$  
    for the Inner, Middle, Outer, and ``Other'' regions of
    the main belt, and for their total.  We assume that they are
    complete for $H\le17,$ and extrapolate to fainter sources using
    the functional form from \eqq{eq:lsstmba} shown as the dotted
    curve.}
  \label{fig:counts}
\end{figure}
We also require the $H$ distribution of potential deflectors.
Figure~\ref{fig:counts} plots our assumptions for $dN/dH.$  The MPC
catalog is likely close to complete in all three regions for $H\le17,$
so we will use the MPC counts directly in this regime.  Note that the
three regions have similar shapes of $dN/dH.$  For $H>17,$ we adopt
the functional form for the cumulative MBA counts given by
initial LSST predictions \citep{LSST_science_book}:
\begin{equation}
  N(<H) \propto \frac{10^{0.43(H-15.7)}}{10^{0.18(H-15.7)} +
    10^{-0.18(H-15.7)}},
\label{eq:lsstmba}
\end{equation}
normalizing this curve to the observed $H<17$ counts for each region.

Our integral for the uncertainty due to unmodelled MBA gravitation
requires the RMS error $\sigma_M(H)$ between the true mass $M$ and the
mass used (if any) in the ephemeris model.
The two largest asteroids, Ceres and Vesta, have masses determined to
high precision by the \textit{Dawn} spacecraft.  Thus despite holding
half of the mass of the asteroid belt, their $\sigma_M$ values are just
$\approx3\times10^{-15} M_\odot$ and
the unknown part of their contributions to other MBA's motion are
quite small.  Aside from a handful of other asteroids in
well-characterized binary systems or visited by spacecraft, most other knowledge of MBA masses
has and will come from mutual gravitational encounters.   An
individual asteroid's mass can be estimated from its measured effect  on the
few MBAs on which it imparts the largest impulses.  The accuracy
$\sigma_M$ of this determination will depend upon the geometry of
these encounters, the measurement errors on the tracers, and the size
of the noise from unmodelled mass that we aim to determine.  These
factors are independent of the mass of the deflector.  We will
assign a value $\sigma_M$ to the RMS uncertainty in MBA mass
attainable from following a few individual tracers, and calculate
the Brownian-motion noise as a function of $\sigma_M.$  We take
\begin{equation}
  \sigma_M(H) = \textrm{min}\left[ \sigma_M, M(H) \right],
\end{equation}
because a given MBA could have a mass determination accurate to
$\sigma_M$ if included in the ephemeris model, or if $M(H)<\sigma_M,$
it would be omitted from the ephemeris model.

To give some context, the most recent
comprehensive investigation of MBA masses from mutual scattering
events was by \citet{baer}, who report typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs.  We can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry.

@@

\subsubsection{Results}

@@

\newpage
\bibliographystyle{aasjournal}
%\bibliography{references}

>>>>

\newcommand{\matA}{A}
\newcommand{\matB}{B}
\newcommand{\matC}{C}

\newcommand{\nIr}{\left\langle nI_r^2\right\rangle}
\newcommand{\nIphi}{\left\langle nI_\phi^2\right\rangle}
\newcommand{\nIz}{\left\langle nI_z^2\right\rangle}

\newcommand{\Var}{\textrm{Var}}
\newcommand{\vcirc}{v_c}
\newcommand{\vrel}{v_{\rm r}}
\newcommand{\Msun}{M_\odot}
\newcommand{\covm}{C}
\newcommand{\lop}{\varpi}
\newcommand{\Lhat}{\hat L}
\usepackage{natbib} 
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage{xcolor}
\usepackage{float}
\usepackage{hyperref}
%\usepackage{lineno}
%\linenumbers

@@

High-precision measurements of positions of Solar System bodies have
great value to multiple scientific pursuits.  They can reveal
gravitational accelerations due to undiscovered bodies such as a
hypothetical ``Planet X'' \citep[\eg][]{holmanP9,inpopP9,trojans,occultations} or passing primordial
black holes \citep{pbh}, determine
the masses of known bodies \citep[\eg][]{goffin,baer}, test the laws of physics, and characterize 
the non-gravitational forces that drive many forms of planetary
migration, \eg\ the maintainence of the near-Earth asteroid (NEA)
population \citep{yorp}.  While most applications of precision orbit determination
have used the 8 major planets as test particles, there are now $>10^6$
known minor planets available for this task.  These small bodies' astrometric
information content is growing extremely rapidly through large-scale
survey projects that can produce
milliarcsecond-scale accuracy for magnitude-limited small-body populations, such as Gaia \citep{gaiass3},
PanSTARRS,\footnote{\url{https://www2.ifa.hawaii.edu/research/Pan-STARRS.shtml}}
and particularly the upcoming Rubin observatory
\textit{Legacy Survey of Space and Time} (LSST).\footnote{\url{https://rubinobservatory.org/}}  The great majority
of currently-tracked objects are main-belt asteroids (MBAs), so we
wish to investigate the potential power of MBAs as gravitational test
bodies.

The most advanced ephemerides for the Solar System \citep{de440,inpop,pitjeva} consider the
gravitational forces emanating from the Sun, from the major planets and their large
moons, from up to $\approx300$ of the largest individual MBAs, and
from distributed mass annuli representing the smaller members of
 the asteroid belt(s) and Kuiper belt.
Here we ask: \textit{what is the RMS ephemeris error in angular or radial
position a typical MBA accrues because of its gravitational
encounters with other individual MBAs?}  We are interested not in the
perturbations that can be well described by a multipolar ring model
for the full population; rather the Brownian motion that accumulates
from closer encounters with: (a) MBAs that \emph{are not included} in the
ephemeris model at all; and (b) asteroids that \emph{are} included
in an ephemeris model, but have some uncertainty $\sigma_M$ in their masses.
Any such departures became a source of noise in any
inferences that we make from use of these MBAs as
dynamical tracer particles.  We aim to quantify this noise, as a
function of the RMS uncertainty $\sigma_M$ on the masses of individual
asteroids, and assuming that deflectors with $M<\sigma_M$ are not
included in the ephemeris at all.

In principal, one can collect all of the positional information for
all $>10^6$ known solar-system bodies, and fit them to a model with
a free state vector and mass for each body.  Calculating the influence
of $>7\times10^6$ free parameters on $10^8$ or so observables is,
however, unlikely to be computationally feasible, and most of the
bodies' masses will be lost in measurement noise and/or be degenerate
with other bodies' masses.  The most ambitious attempt at a global
solution that we are aware of is that of \citet{goffin}, who was able
to include 250 MBAs as gravitating bodies in the fit to the full
corpus of observations at the time.  For only $\approx 1/2$ of the
bodies did the fit yield masses $>3\times$ the uncertainties.  With
vast improvements in data and computing power since then and in the
next decade, the number of detectable
asteroid masses should increase, but there will always be
uncertainties that accrue from the unmodelled deflectors and the
finite $\sigma_M$ of modelled deflectors.  These remnant uncertainties
will in turn serve as noise in attempts to model new sources of
acceleration.

While long-term diffusion of orbital elements has been extensively
investigated in the context of planetary-system formation and
planetary migration, we address here the more limited question of how
much the MBAs will have their orbit elements and positions altered
by Brownian motion over a time period $T \lesssim100$~yr of human
observation, in the present dynamical environment.  We will make use
of the currently known population of MBAs, correcting for
incompleteness of current surveys at lower masses---or more precisely,
at fainter absolute magnitudes $H,$ since the $H$ distribution is
known far more accurately than the $M$ distribution.

\section{Calculation overview}
We aim to calculate the Brownian motion variance to an accuracy of a
factor $\approx 2.$
Greater accuracy in the calculation is not warranted
since some of the principal inputs have substantial uncertainty.  The
number of MBAs vs mass is 
uncertain because of unknowns in the conversion
from $H$ to mass; the status of future observations and $\sigma_M$ is
also unclear.  Hence we can determine our quantities only to leading
order in the MBA orbital eccentricity $e$ of the tracer population.
We will assume throughout that the joint
distribution of mass (or $H$) and orbital elements is separable into a
mass distribution and an orbital distribution.  In this scenario, the
Brownian motion is independent of the tracer mass, and involves an
integral over the mass distribution of the deflecting bodies.

@@

All distances will be given in units of the original semimajor axis $a_0$, and all velocities in units of the circular velocity $\vcirc \equiv \sqrt{G\Msun/a_0}.$  In these units, $G\Msun=1,$ the period of the initial orbit is $2\pi,$ and the (unperturbed) mean anomaly is $M=t+M_0.$

@@

One of our tasks will be to derive, from the known asteroid population, the rate (per target) of encounters vs the imparted impulse,
\begin{equation}
  \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z},
  \label{eq:dN}
\end{equation}
where the components of $\vecI$ are given in the cylindrical basis
vectors about the asteroid's position at the impulse.
We will estimate this function from the known population,
approximating it as constant within each of three subsets of tracer MBAs:
the inner, middle, and outer regions bounded by the 4:1, 3:1,
5:2, and 2:1 mean-motion resonances with Jupiter.
We will further assume that the impulses on a given target are drawn
independently from this distribution, \ie\ a Poisson process defined
by this rate.  In this case, the only properties of the impulse
distribution that we need are its second moments for components $\alpha \in \{r,\phi,z\}$
\begin{equation}
  \left \langle n I_\alpha^2 \right\rangle \equiv \int d^3I \frac{dN}{dt\,dI_r\,dI_\phi\,dI_z} I_\alpha^2.
\label{eq:nvsq}
\end{equation}
The average of the cross terms $I_rI_z$ and $I_\phi I_z$ vanishes if
the deflector distribution is symmetric in inclination as expected, and we find numerically that the mean $I_rI_\phi$ is small enough to ignore.

From this knowledge of the impulse distribution, our goal is to obtain
the covariance matrix of the deviations of the target's position from
the initial orbit, after some time $T.$ The position observables are
the range, plus the celestial latitude and longitude of the target.  We will simplify our results by assuming a heliocentric observer, so that the observational position vector is $\vecp\equiv (r,\theta=z/r,\phi).$ The quantity we seek is the covariance matrix $\covm^p$ of the observations attributable to the accumulated gravitational perturbations:
\begin{equation}
  \covm^p  \equiv  \left\langle  \Delta\vecp \Delta\vecp^T\right\rangle,  \label{eq:Cp}
\end{equation}
where the angle brackets indicate an average over possible
realizations of the impulse history, and over the mean anomaly $M$ at the time of observation.  To do so, we will introduce an intermediate set of 6 parameters $\vecq$ describing the deviations of the orbital elements induced by the impulses.  The $\vecq$ components will be selected to be equal to zero before any impulses are applied, and each responds linearly to $\vecI$ at $|I|\ll 1.$
We will derive the matrix $\matA$ that describes the orbital-element shifts at time $T$ that arise from an impulse at time $t_i$: 
\begin{equation}
  \Delta \vecq(T,t_i) = \matA(e_0, T, t_i) \cdot \vecI.
  \label{eq:A}
\end{equation}
In our first-order perturbation theory, $\vecq(T)$ will be the sum of the $\Delta\vecq_i$ imparted by all impulses applied at times $0<t_i<T.$  Because the impulses are uncorrelated, $\vecq$ will therefore be the result of a random walk.  The distribution will have a covariance matrix $\covm^q(T) \equiv \left\langle \vecq(T) \vecq^T(T) \right\rangle$ whose elements are
\begin{eqnarray}
  \covm^q_{jk}(e_o,T) & = & \left\langle \sum_{i,\gamma} \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) I^2_{i,\gamma} \right\rangle \\
           & = & \sum_\gamma \int dt_i \matA_{j\gamma}(e_0,T,t_i) \matA_{k\gamma}(e_0,T,t_i) \left\langle n I_\gamma^2\right\rangle.
\label{eq:Cqjk}
\end{eqnarray}
In the first line, the sum $i$ runs over the impulses and $\gamma$ runs over the components $r,\phi,z$ of the impulse. The second line evaluates the expectation value of averaging over realizations of the random walk of impulses, exploiting the independence of the individual impulses from each other.

The last element of our calculation will be a conversion from the
element shifts $\vecq$ into the observed displacements $\Delta\vecp$
at time $T.$  In linear perturbation theory this will again be expressible as a matrix
\begin{equation}
  \Delta\vecp(e_0,T) = \matB(e_0,T) \cdot \vecq(T).
\label{eq:B}
\end{equation}
Combining this with \eqq{eq:Cp} and \eqq{eq:Cqjk} yields the desired result
\begin{eqnarray}
  \covm^p_{\alpha\beta}(e_0,T) & = & \sum_{jk} B_{\alpha j}(e_0,T) B_{\beta k}(e_0,T) \covm^q_{jk}(T) \nonumber \\
  & = & \sum_\gamma \left\langle nI_\gamma^2\right\rangle \int dt_i \left[AB(e_0,T,t_i)\right]_{\alpha\gamma}  \left[AB(e_0,T,t_i)\right]_{\beta\gamma} 
\label{eq:ABq}
\end{eqnarray}
which we would wish to average over the phase of the initial mean anomaly $M_0$ of the MBA.

Section~\ref{sec:impulse} describes the estimation of $\left\langle
  nI_\alpha^2\right\rangle$ for MBA regions.
Section~\ref{sec:propagation} derives the forms of $\matA(e_0,T,t_i),$
and $\matB(e_0,T).$  Section~\ref{sec:results} combines these and
summarizes the results and their implications.

\section{Impulse distribution}
\label{sec:impulse}
\subsection{Derivation}
Under the assumption that the distributions of masses and orbital
elements of MBAs are separable, the quantities needed to describe the
random walk of the asteroid's orbits can be expressed as an integral
over the absolute magnitude $H$ of the deflectors, the velocity $v$,
impact parameter $b,$ and polar/azimuthal angles $\theta,\phi$ of the
unit vector of the impulse direction $\bhat$:
\begin{eqnarray}
  \left\langle n I_\gamma^2 \right\rangle & = & \int
                                              dH\,dv\,db\,d\theta\,d\phi
                                             \frac{dN}{dH} \frac{dn}{dv\,db\,d\theta\,d\phi}
                                              I^2_\gamma(H,v,b,\theta\phi)
  \\
  & = & \int dH \frac{dN}{dH} \int dv
        \frac{dn}{dA\,dv\,d\theta\,d\phi} \hat b_\gamma^2(\theta,\phi)
        \int_{b_{\rm min}}^{b_{\rm max}} 2\pi b\,db\,
        \left(\frac{\sigma_M(H)}{bv}\right)^2 \\
  & = & 2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int dv
        \frac{dn}{dA\,dv} v^{-2} \left\langle\hat b_\gamma^2(v)\right\rangle
        \log(b_{\rm max}/b_{\rm min}).
\end{eqnarray}
The second row adopts the (numerically verified) assumption that the
distribution of impact parameter $b$ will always be $\propto dA=2\pi
b\,db$ in the range of $b$ of interest to us.  The mean rate of encounters
between a deflector-tracer pair, as a function of the impact velocity
and geometry, is $dn/dA\,dv\,d\theta\,d\phi,$ which we will determine
through numerical measurements with the orbits of all known MBAs. The
$H$ distribution of MBAs is $dN/dH.$ We denote as $\sigma_M(H)$
the RMS error in the mass of MBAs at $H$ from the value used in the ephemeris model.
We use $\sigma_M$ rather than $M$ since we are seeking the size of
deviations from the ephemeris model.  For MBAs that are not included
in the ephemeris model, $\sigma_M=M(H),$ the full mass of the deflector.

@@

For $b_{\rm min},$ we adopt the criterion that encounters at
sufficiently small $b$ will generate sufficiently large impulses $I$
that observations of the tracer will detect this individual impulse at
levels well above measurement noise.  This would mean that an
ephemeris model could include the mass of the deflector asteroid, and
this mass could be determined from fitting the data of the tracer.
Such encounters would thus no longer be considered source of
stochastic Brownian motion.  Finding all such encounters among known
MBAs is feasible, and in fact is forecasted for the next decade's LSST
observations by
\citet{negin}. The fitting of the tracer data for the mass of the
deflector is also a small computational effort.

If we crudely choose some threshold
$I_{\rm det}$ of impulse as being large enough to generate high-$S/N$
deflections on its tracer, then our condition becomes $M(H)/bv <
I_{\rm det},$ or $b_{\rm min}=M(H)/vI_{\rm det}.$  We then have
\begin{equation}
   \left\langle n I_\gamma^2 \right\rangle = 
2\pi \int dH \frac{dN}{dH} \sigma^2_M(H)  \int_{\sqrt{M(H)/I_{\rm
      det}}}^\infty \frac{dv}{v^2} 
        \frac{dn}{dA\,dv}  \left\langle\hat b_\gamma^2(v)\right\rangle
        \log \left[v^2 I_{\rm det}/M(H)\right].
        \label{eq:nI2}
      \end{equation}
The lower bound on the $v$ integral marks the speed below which
$b_{\rm max}<b_{\rm min}.$  Since $b_{\rm min}$ and $b_{\rm max}$
appear logarithmically, the estimated diffusion is relatively insensitive to choices of
their values.

\subsection{Numerical results}
\eqq{eq:nI2} reduces the calculation to a double integral over the
deflector $H$ and the tracer-deflector relative velocity $v.$  The
constituents of this calculation are estimated as follows.

@@

Using a simple leapfrog integrator with gravity from the Sun and the 8
major planets, we advance all MBA's orbits from their heliocentric
osculating elements at the epochs given by the MPC, to barycentric
state vectors on 1 May 2025.  Using $kd$-trees to accelerate
pair-finding, we locate all tracer-deflector pairs
that pass within 0.03~AU of each other during the $\pm1$-day period
around this initial epoch, assuming inertial relative motion during
this interval.  The circumstances of such encounters are saved: the
identities of the two MBAs involved, the time of closest approach, the
relative velocity $v,$ and the impact parameter vector $\vecb.$

@@

\subsubsection{Asteroid properties}
We crudely approximate the population as having a geometric albedo of 0.25,
density of 2500~kg~m$^{-3},$ and spherical shapes, which leads to the
conversion
\begin{equation}
  M(H) = 1.2\times10^{-17} M_\odot \times 10^{-0.6(H-15)}
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{mbacounts.pdf}
  \caption{Differential counts of MBAs cataloged by the MPC are
    plotted vs absolute magnitude $H$  
    for the Inner, Middle, Outer, and ``Other'' regions of
    the main belt, and for their total.  We assume that they are
    complete for $H\le17,$ and extrapolate to fainter sources using
    the functional form from \eqq{eq:lsstmba} shown as the dotted
    curve.}
  \label{fig:counts}
\end{figure}
We also require the $H$ distribution of potential deflectors.
Figure~\ref{fig:counts} plots our assumptions for $dN/dH.$  The MPC
catalog is likely close to complete in all three regions for $H\le17,$
so we will use the MPC counts directly in this regime.  Note that the
three regions have similar shapes of $dN/dH.$  For $H>17,$ we adopt
the functional form for the cumulative MBA counts given by
initial LSST predictions \citep{lsstbook}:
\begin{equation}
  N(<H) \propto \frac{10^{0.43(H-15.7)}}{10^{0.18(H-15.7)} +
    10^{-0.18(H-15.7)}},
\label{eq:lsstmba}
\end{equation}
normalizing this curve to the observed $H<17$ counts for each region.

Our integral for the uncertainty due to unmodelled MBA gravitation
requires the RMS error $\sigma_M(H)$ between the true mass $M$ and the
mass used (if any) in the ephemeris model.
The two largest asteroids, Ceres and Vesta, have masses determined to
high precision by the \textit{Dawn} spacecraft \citep{Ceres,Vesta}.  Thus despite holding
half of the $\approx 10^{-9} M_\odot$ mass of the asteroid belt, their $\sigma_M$
values\footnote{More precisely, the uncertainties in their $GM$ values
  relative to $GM_\odot$.} are 
$\le3\times10^{-15} M_\odot$ and
% Konilov papers: 1.2e-5 km^3/s^2 for Vesta, 4e-4 for Ceres
this unknown part of their contribution to other MBA's motion is 
unimportant.  Aside from a handful of other asteroids in
well-characterized binary systems or visited by spacecraft, most other knowledge of MBA masses
has and will come from mutual gravitational encounters.   An
individual asteroid's mass can be estimated from its measured effect  on the
few MBAs on which it imparts the largest impulses.  The accuracy
$\sigma_M$ of this determination will vary, depending upon the
geometry  and timing of
each deflector MBA's encounters, and the measurement errors on its tracers.  These
factors are independent of the mass of the deflector.  We will
assign a value $\sigma_M$ to the RMS uncertainty in MBA mass
attainable from following a few individual tracers, and calculate
the Brownian-motion noise as a function of $\sigma_M.$  We take
\begin{equation}
  \sigma_M(H) = \textrm{min}\left[ \sigma_M, M(H) \right],
\end{equation}
because a given MBA could have a mass determination accurate to
$\sigma_M$ if included in the ephemeris model, or if $M(H)<\sigma_M,$
it would be omitted from the ephemeris model.

To give some context, a
comprehensive investigation of MBA masses from mutual scattering
events by \citet{goffin} reports typical $\sigma_M\approx
5\times10^{-13}$ for 30 MBAs.  We can expect substantial improvement
in the LSST era when a millions of potential tracer asteroids attain
mas-level astrometry.

@@

@@

\newpage
\bibliographystyle{aasjournal}
\bibliography{references}
